Performance analysis of high-resolution \\ ice sheet simulations

### 2022-11-06 paper/perfmod.tex
clean .tex for submission

\usepackage{verbatim,xspace,amsmath,amssymb,bm,multirow,dashbox}
\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{igsnatbib,lineno}

\usepackage{hyperref}
\hypersetup{pdfauthor={Ed Bueler},
            pdfcreator={pdflatex},
            colorlinks=true,
            citecolor=purple,
            linkcolor=red,
            urlcolor=blue,
            }

% math macros
\newcommand\bb{\mathbf{b}}
\newcommand\bc{\mathbf{c}}
\newcommand\bbf{\mathbf{f}}
\newcommand\bg{\mathbf{g}}
\newcommand\bn{\mathbf{n}}
\newcommand\bq{\mathbf{q}}
\newcommand\bu{\mathbf{u}}
\newcommand\bv{\mathbf{v}}
\newcommand\bx{\mathbf{x}}
\newcommand\by{\mathbf{y}}

\newcommand\bA{\mathbf{A}}
\newcommand\bF{\mathbf{F}}
\newcommand\bH{\mathbf{H}}
\newcommand\bN{\mathbf{N}}
\newcommand\bQ{\mathbf{Q}}
\newcommand\bU{\mathbf{U}}
\newcommand\bV{\mathbf{V}}
\newcommand\bW{\mathbf{W}}
\newcommand\bX{\mathbf{X}}

\newcommand\bzero{\bm{0}}

\newcommand{\Div}{\nabla\cdot}
\newcommand\eps{\epsilon}
\newcommand{\grad}{\nabla}
\newcommand{\ip}[2]{\ensuremath{\left<#1,#2\right>}}
\newcommand\lam{\lambda}
\newcommand\lap{\triangle}
\newcommand\RR{\mathbb{R}}
\newcommand{\half}{\tfrac{1}{2}}

\newcommand{\Divx}{\nabla_\bx \cdot}
\newcommand{\gradx}{\nabla_\bx}

\newcommand{\rhoi}{\rho_{\text{i}}}
\newcommand{\pp}{{\text{p}}}
\newcommand{\qq}{{\text{q}}}
\newcommand{\rr}{{\text{r}}}

\newcommand{\mR}{R^{\bm{\oplus}}}
\newcommand{\iR}{R^{\bullet}}

\newcommand{\sold}{s_{\text{o}}}
\newcommand{\told}{t_{\text{o}}}

\newcommand{\byold}{\by_{\text{o}}}

\newcommand{\onecol}[1]{\includegraphics[width=86mm]{#1}}
\newcommand{\onecolless}[1]{\includegraphics[width=80mm]{#1}}
\newcommand{\onecolthin}[1]{\includegraphics[width=75mm]{#1}}

%\newcommand{\twocol}[1]{\includegraphics[width=178mm]{#1}}
\newcommand{\twocol}[1]{\includegraphics[width=165mm]{#1}}
\newcommand{\twocolless}[1]{\includegraphics[width=140mm]{#1}}

\newcommand{\ds}{\displaystyle}

@@

%\keywords{glacier flow,glacier modelling,ice-sheet modelling}
% suggested reviewers:  Dan Goldberg, William Lipscomb, Toby Isaac, Josefin Ahlkrona, Alex Jarosch, Doug Brinkerhoff

@@

\newcommand{\nn}{\text{n}}
A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, for the isothermal case where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}, one has $C = 2 A (\rho g)^\nn/(\nn+2)$.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.

@@

%         References
\bibliography{perfmod}
\bibliographystyle{igs}

>>>>

\usepackage{amsmath}
\usepackage{igsnatbib}
\usepackage{lineno}

\usepackage{hyperref}
\hypersetup{pdfauthor={Ed Bueler},
            pdfcreator={pdflatex},
            colorlinks=true,
            citecolor=blue,
            linkcolor=red,
            urlcolor=blue,
            }

% math macros
\newcommand\bu{\mathbf{u}}
\newcommand\bx{\mathbf{x}}
\newcommand{\Divx}{\nabla_\bx \cdot}
\newcommand{\gradx}{\nabla_\bx}
\newcommand{\nn}{\text{n}}

@@

@@

A diffusive description of the mass continuity equation is also valid in the small-aspect-ratio limit which generates the SIA \citep{SchoofHewitt2013}:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(d\, \gradx s \right) + a. \label{eq:siamasscontinuity}
\end{equation}
Here $d = C H^{\nn+2} |\gradx s|^{\nn-1}$ is the nonlinear diffusivity.\footnote{In detail, for the isothermal case where $A$ is the ice softness, $\rho$ is the ice density, $g$ is gravity, and $\nn\approx 3$ is the Glen exponent in the flow law \citep{GreveBlatter2009}, one has $C = 2 A (\rho g)^\nn/(\nn+2)$.}  While Equation \eqref{eq:siamasscontinuity} does not hold directly for Stokes or other membrane-stress-resolving dynamics, the same diffusivity $d$, an essentially geometric quantity, can be computed regardless.  Generically across stress balance choices, for grounded glaciers and ice sheets one observes that large values of $d$ indicate locations of unstable mode growth if explicit time-steps are chosen too large.

@@

%         References
%\bibliography{perfmod}
%\bibliographystyle{igs}

\begin{thebibliography}{34}
\providecommand{\natexlab}[1]{#1}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[\protect\citename{Aschwanden and others, }2019]{Aschwandenetal2019}
Aschwanden A, Fahnestock MA, Truffer M, Brinkerhoff DJ, Hock R, Khroulev C,
  Mottram R and Khan SA (2019) Contribution of the {Greenland Ice Sheet} to sea
  level over the next millennium. \emph{Science Advances}, \textbf{5}(6) (\doi
  {10.1126/sciadv.aav9396})

\bibitem[\protect\citename{Briggs and others, }2000]{Briggsetal2000}
Briggs W, Henson VE and McCormick S (2000) \emph{A {M}ultigrid {T}utorial}.
  SIAM Press, Philadelphia, 2nd edition

\bibitem[\protect\citename{Brinkerhoff and Johnson,
  }2015]{BrinkerhoffJohnson2015}
Brinkerhoff DJ and Johnson JV (2015) Dynamics of thermally induced ice streams
  simulated with a higher-order flow model. \emph{Journal of Geophysical
  Research: Earth Surface}, \textbf{120}(9), 1743--1770 (\doi
  {10.1002/2015JF003499})

\bibitem[\protect\citename{Brown and others, }2013]{BrownSmithAhmadia2013}
Brown J, Smith B and Ahmadia A (2013) Achieving textbook multigrid efficiency
  for hydrostatic ice sheet flow. \emph{SIAM Journal on Scientific Computing},
  \textbf{35}(2), 359--375 (\doi {10.1137/110834512})

\bibitem[\protect\citename{Bueler, }2016]{Bueler2016}
Bueler E (2016) Stable finite volume element schemes for the shallow ice
  approximation. \emph{Journal of Glaciology}, \textbf{62}(232), 230--242 (\doi
  {10.1017/jog.2015.3})

\bibitem[\protect\citename{Bueler, }2021{\natexlab{a}}]{Bueler2021conservation}
Bueler E (2021{\natexlab{a}}) Conservation laws for free-boundary fluid layers.
  \emph{SIAM Journal on Applied Mathematics}, \textbf{81}(5), 2007--2032 (\doi
  {10.1137/20M135217X})

\bibitem[\protect\citename{Bueler, }2021{\natexlab{b}}]{Bueler2021}
Bueler E (2021{\natexlab{b}}) \emph{{PETSc} for {P}artial {D}ifferential
  {E}quations: {N}umerical {S}olutions in {C} and {P}ython}. SIAM Press,
  Philadelphia

\bibitem[\protect\citename{Bueler and others, }2005]{Bueleretal2005}
Bueler E, Lingle CS, Kallen-Brown JA, Covey DN and Bowman LN (2005) Exact
  solutions and verification of numerical models for isothermal ice sheets.
  \emph{Journal of Glaciology}, \textbf{51}(173), 291--306 (\doi
  {10.3189/172756505781829449})

\bibitem[\protect\citename{Calvo and others, }2003]{Calvoetal2003}
Calvo N and others (2003) On a doubly nonlinear parabolic obstacle problem
  modelling ice sheet dynamics. \emph{SIAM Journal on Applied Mathematics},
  \textbf{63}(2), 683--707 (\doi {10.1137/S0036139901385345})

\bibitem[\protect\citename{Cheng and others, }2017]{Chengetal2017}
Cheng G, L{\"o}tstedt P and von Sydow L (2017) Accurate and stable time
  stepping in ice sheet modeling. \emph{Journal of Computational Physics},
  \textbf{329}, 29--47 (\doi {10.1016/j.jcp.2016.10.060})

\bibitem[\protect\citename{Clarke and others, }2015]{Clarkeetal2015}
Clarke GKC, Jarosch AH, Anslow FS, Radić V and Menounos B (2015) Projected
  deglaciation of western {C}anada in the twenty-first century. \emph{Nature
  Geoscience}, \textbf{8}, 372--377 (\doi {10.1038/ngeo2407})

\bibitem[\protect\citename{Fischler and others, }2022]{Fischleretal2022}
Fischler Y, R\"uckamp M, Bischof C, Aizinger V, Morlighem M and Humbert A
  (2022) A scalability study of the {Ice-sheet and Sea-level System Model}
  ({ISSM}, version 4.18). \emph{Geoscientific Model Development},
  \textbf{15}(9), 3753--3771 (\doi {10.5194/gmd-15-3753-2022})

\bibitem[\protect\citename{Gladstone and others, }2017]{Gladstoneetal2017}
Gladstone RM, Warner RC, Galton-Fenzi BK, Gagliardini O, Zwinger T and Greve R
  (2017) Marine ice sheet model performance depends on basal sliding physics
  and sub-shelf melting. \emph{The Cryosphere}, \textbf{11}(1), 319--329 (\doi
  {10.5194/tc-11-319-2017})

\bibitem[\protect\citename{Goelzer and others, }2020]{Goelzeretal2020}
Goelzer H, Nowicki S and others (2020) The future sea-level contribution of the
  {G}reenland ice sheet: a multi-model ensemble study of {ISMIP6}. \emph{The
  Cryosphere}, \textbf{14}(9), 3071--3096 (\doi {10.5194/tc-14-3071-2020})

\bibitem[\protect\citename{Greve and Blatter, }2009]{GreveBlatter2009}
Greve R and Blatter H (2009) \emph{Dynamics of {I}ce {S}heets and {G}laciers}.
  Advances in Geophysical and Environmental Mechanics and Mathematics,
  Springer, Berlin

\bibitem[\protect\citename{Gudmundsson, }2013]{Gudmundsson2013}
Gudmundsson GH (2013) Ice-shelf buttressing and the stability of marine ice
  sheets. \emph{The Cryosphere}, \textbf{7}(2), 647--655 (\doi
  {10.5194/tc-7-647-2013})

\bibitem[\protect\citename{Hindmarsh and Payne, }1996]{HindmarshPayne1996}
Hindmarsh RCA and Payne AJ (1996) Time-step limits for stable solutions of the
  ice-sheet equation. \emph{Annals of Glaciology}, \textbf{23}, 74--85 (\doi
  {10.3189/S0260305500013288})

\bibitem[\protect\citename{Hock, }2005]{Hock2005}
Hock R (2005) Glacier melt: a review of processes and their modelling.
  \emph{Progress in Physical Geography}, \textbf{29}(3), 362--391 (\doi
  {10.1191/0309133305pp453ra})

\bibitem[\protect\citename{Hoffman and others, }2018]{Hoffmanetal2018}
Hoffman MJ, Perego M, Price SF, Lipscomb WH, Zhang T, Jacobsen D, Tezaur I,
  Salinger AG, Tuminaro R and Bertagna L (2018) {MPAS-Albany Land Ice (MALI)}:
  a variable-resolution ice sheet model for {E}arth system modeling using
  {V}oronoi grids. \emph{Geoscientific Model Development}, \textbf{11}(9),
  3747--3780 (\doi {10.5194/gmd-11-3747-2018})

\bibitem[\protect\citename{Isaac and others, }2015]{IsaacStadlerGhattas2015}
Isaac T, Stadler G and Ghattas O (2015) Solution of nonlinear {S}tokes
  equations discretized by high-order finite elements on nonconforming and
  anisotropic meshes, with application to ice sheet dynamics. \emph{SIAM
  Journal on Scientific Computing}, \textbf{37}(6), B804--B833 (\doi
  {10.1137/140974407})

\bibitem[\protect\citename{Jouvet and Bueler, }2012]{JouvetBueler2012}
Jouvet G and Bueler E (2012) Steady, shallow ice sheets as obstacle problems:
  well-posedness and finite element approximation. \emph{SIAM Journal on
  Applied Mathematics}, \textbf{72}(4), 1292--1314 (\doi {10.1137/110856654})

\bibitem[\protect\citename{Jouvet and Gr{\"a}ser, }2013]{JouvetGraeser2013}
Jouvet G and Gr{\"a}ser C (2013) An adaptive {N}ewton multigrid method for a
  model of marine ice sheets. \emph{Journal of Computational Physics},
  \textbf{252}, 419--437 (\doi {10.1016/j.jcp.2013.06.032})

\bibitem[\protect\citename{Leng and others, }2012]{Lengetal2012}
Leng W, Ju L, Gunzburger M, Price S and Ringler T (2012) A parallel high-order
  accurate finite element nonlinear {S}tokes ice sheet model and benchmark
  experiments. \emph{Journal of Geophysical Research: Earth Surface},
  \textbf{117}(F1) (\doi {10.1029/2011JF001962})

\bibitem[\protect\citename{LeVeque, }2007]{LeVeque2007}
LeVeque RJ (2007) \emph{{Finite Difference Methods for Ordinary and Partial
  Differential Equations: Steady-State and Time-Dependent Problems}}. SIAM
  Press, Philadelphia

\bibitem[\protect\citename{L{\"o}fgren and others,
  }2022]{LofgrenAhlkronaHelanow2022}
L{\"o}fgren A, Ahlkrona J and Helanow C (2022) Increasing stable time-step
  sizes of the free-surface problem arising in ice-sheet simulations.
  \emph{Journal of Computational Physics: X}, \textbf{16}(100114) (\doi
  {10.1016/j.jcpx.2022.100114})

\bibitem[\protect\citename{Robinson and others, }2022]{Robinsonetal2022}
Robinson A, Goldberg D and Lipscomb WH (2022) A comparison of the stability and
  performance of depth-integrated ice-dynamics solvers. \emph{The Cryosphere},
  \textbf{16}(2), 689--709 (\doi {10.5194/tc-16-689-2022})

\bibitem[\protect\citename{Schoof and Hewitt, }2013]{SchoofHewitt2013}
Schoof C and Hewitt IJ (2013) Ice-sheet dynamics. \emph{Annual Review of Fluid
  Mechanics}, \textbf{45}, 217--239 (\doi {10.1002/jgrf.20146})

\bibitem[\protect\citename{Seguinot and Delaney, }2021]{SeguinotDelaney2021}
Seguinot J and Delaney I (2021) Last-glacial-cycle glacier erosion potential in
  the {A}lps. \emph{Earth Surface Dynamics}, \textbf{9}(4), 923--935 (\doi
  {10.5194/esurf-9-923-2021})

\bibitem[\protect\citename{Seroussi and others, }2020]{Seroussietal2020}
Seroussi H, Nowicki S and others (2020) {ISMIP6 Antarctica}: a multi-model
  ensemble of the {A}ntarctic ice sheet evolution over the 21st century.
  \emph{The Cryosphere}, \textbf{14}(9), 3033--3070 (\doi
  {10.5194/tc-14-3033-2020})

\bibitem[\protect\citename{Trottenberg and others, }2001]{Trottenbergetal2001}
Trottenberg U, Oosterlee CW and Schuller A (2001) \emph{Multigrid}. Elsevier,
  Oxford

\bibitem[\protect\citename{Tuminaro and others, }2016]{Tuminaroetal2016}
Tuminaro R, Perego M, Tezaur I, Salinger A and Price S (2016) A matrix
  dependent/algebraic multigrid approach for extruded meshes with applications
  to ice sheet modeling. \emph{SIAM Journal on Scientific Computing},
  \textbf{38}(5), C504--C532 (\doi {10.1137/15M1040839})

\bibitem[\protect\citename{Weber and others, }2021]{Weberetal2021}
Weber ME, Golledge NR, Fogwill CJ, Turney CSM and Thomas ZA (2021)
  Decadal-scale onset and termination of {A}ntarctic ice-mass loss during the
  last deglaciation. \emph{Nature Communications}, \textbf{12}(6683) (\doi
  {10.1038/s41467-021-27053-6})

\bibitem[\protect\citename{Winkelmann and others, }2011]{Winkelmannetal2011}
Winkelmann R, Martin MA, Haseloff M, Albrecht T, Bueler E, Khroulev C and
  Levermann A (2011) {The Potsdam Parallel Ice Sheet Model (PISM-PIK)--Part 1:
  Model description}. \emph{The Cryosphere}, \textbf{5}(3), 715--726 (\doi
  {10.5194/tc-5-715-2011})

\bibitem[\protect\citename{Wirbel and Jarosch, }2020]{WirbelJarosch2020}
Wirbel A and Jarosch AH (2020) Inequality-constrained free-surface evolution in
  a full {S}tokes ice flow model (\textit{evolve\_glacier v1.1}).
  \emph{Geoscientific Model Development}, \textbf{13}(12), 6425--6445 (\doi
  {10.5194/gmd-13-6425-2020})

\end{thebibliography}
